---
title: "Bolognese Vowel Analysis"
date: "2026-09-04"
date-modified: last-modified
author: 
  - name: Mykel Loren Brinkerhoff
    orcid: 0000-0002-2190-6544
    email: mlbrinke@uw.edu
    affilitation: 
      - name: University of Washington 
        city: Seattle
        state: WA
        url: www.uw.edu
format: 
  html: 
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    toc-location: left-body
    html-math-method: katex
    embed-resources: true
    # number-sections: true
    # number-depth: 3
    code-fold: true
    code-links: 
      - text: GitHub Repository
        repo: https://github.com/mylbrinkerhoff/icphs2027_bolognese
bibliography: bolognese.bib
csl: unified-style-sheet-for-linguistics.csl
---

```{r}
#| echo: false
#| label: package_install
#| message: false 

remotes::install_github("joeystanley/joeyr") # helper functions from joeystanley
remotes::install_github("joeystanley/joeysvowels") # vowel datasets

# renv::install(c("knitr","rmarkdown","fwb"))
### restore R session and packages to versions found in the renv.lock file
# renv::restore()

### Load helper packages
library(dplyr) # for data manipulation, graphic, and data wrangling
library(readr) # loading in data
library(stringr) # look for stuff in strings
library(purrr)
library(knitr)
library(rmarkdown)
library(forcats)
library(reticulate)
library(ggplot2) # for plotting
library(geomtextpath)
library(here) # for creating pathways relative to the top-level directory
library(remotes) # allows accessing github
library(joeyr) # needed for the find_outliers function
library(ggokabeito) # colorblind friendly color based on Okabe-Ito scheme
library(viridis)
library(fwb)
library(sp) # needed for Bhattacharyya's Affinity calculation
library(adehabitatHR) # needed for Bhattacharyya's Affinity calculation
library(boot) # needed for bootstraping

### Load vowel specific packages
library(tidynorm) # package for vowel normalization

```

```{r}
#| echo: false
#| label: helper_functions
#| messages: false

cor_formant <- function(df, F1, F2) {
  
  # Correlation between F1 and F2
  cor <- stats::cor(
    df$F1,
    df$F2,
    use = "complete.obs"
  )

}

# Formula for finding the expected BA for samples of equal and unequal size based on Bernardi (2026).

expected_BA <- function(df, vowel1, vowel2) {
  
  # Find the sample size 
  n1 <- sum(df$phoneme == vowel1)
  n2 <- sum(df$phoneme == vowel2)

  #calculate the mean of the samples
  m <- mean(n1,n2)

  # Calculate expected BA for equal sized samples
  BA_even <- 0.142120 + 0.527779 * atan(0.227416 * m)

  # Check if the sample are equal in size and return BA_even if TRUE and calculate residual and add that to BA_even if FALSE
  if (n1==n2) {

    return(BA_even)    
  
  } else {

    #find the imbalance ratio between the smallest and the largest
    ratio <- min(n1, n2)/max(n1, n2)

    #find the absolute imbalance between n1, n2
    abs_imbalance <- abs(n1-n2)

    #calculate residuals
    residual <- -0.06822 + (0.06197 * ratio) - (0.000547 * abs_imbalance) + (0.002892 * ratio *  abs_imbalance)

    # Add the residuals to BA_even for expected BA for uneven
    BA_uneven <- BA_even + residual

    return(BA_uneven)
  }
}

# Vowel overlap function from Joey Stanley at BYU
# at https://joeystanley.com/blog/vowel-overlap-in-r-advanced-topics/
overlap <- function(F1, F2, vowel, method = "pillai") {
  vowel_data <- data.frame(vowel) %>%
    droplevels()

  if (nrow(vowel_data) < 1) {
    return(NA)
  } # not zero
  if (length(table(vowel_data)) < 2) {
    return(NA)
  } # both vowels are represented
  if (min(table(vowel_data)) < 5) {
    return(NA)
  } # there are at least 5 of each vowel

  if (method == "pillai") {
    summary(manova(cbind(F1, F2) ~ vowel))$stats["vowel", "Pillai"]
  } else if (method == "BA") {
    adehabitatHR::kerneloverlap(
      sp::SpatialPointsDataFrame(cbind(F1, F2), vowel_data),
      method = 'BA'
    )[2, 1]
  } else {
    warning("Improper method")
    return(NA)
  }
}

threshold_BA <- function(formant_cor, BA_expected) {

  return(-0.019907 + (-0.242121 * formant_cor) + (1.017491 * BA_expected) +  (0.277828 * formant_cor * BA_expected))
  
}

```

## Introduction

This page is dedicated to presenting an analysis of Bolognese vowels based on the audio accompanying the lessons found in @vitaliDscdegarretBulgnaisManuale2009's *Dscårret in bulgnaiṡ?: manuale e grammatica del dialetto bolognese*, which is a grammar and lesson manual that introduces the Emilian dialect spoken in Bologna. 

Current descriptions claim that there are three low vowel qualities in the language that are described as /æ/, /a/, and /ɑ/ [@canepariPronunciaGrafiaBolognese1995]. The central question is whether there is sufficient acoustic evidence for these three vowels or if there is evidence for a subset of these vowels. Another issue is that there is some debate about whether /æ/ is [æ] or [ɛ]. In @SonDalBulgnais this vowel is descibed as the short version of [ɛː].

The reason for this is that descriptively these vowels are described as having nonoverlapping environments and generally lack minimal pairs. For example, /æ/ is claimed to only exist before velar nasals and in a few other functional words like demonstratives [see @canepariPronunciaGrafiaBolognese1995; @vitaliDscdegarretBulgnaisManuale2009 for a more detail discussion] and doesn't have minimal pairs with the other two phones. All of this makes phonologists' spider sense tingle that we are dealing with allophones rather than phonemes. 

I will use several tools to explore whether or not this is the case. The two most commonly used tools are Bhattacharrya's Coeffecient/Affinity [BA; @bhattacharyyaMeasureDivergenceTwo1943] and the Pillai-Bartlett trace [Pillai; @pillaiNewTestCriteria1955]. These both have been used quite extensively to invesatigate vowel mergers in sociolinguistic, applied linguistic, and fieldwork research [@hayFactorsInfluencingSpeech2006;@nyczBestPracticesMeasuring2014;@johnsonQuantifyingOverlapBhattacharyyas2015;
@amengualEffectsLanguageDominance2016;
@kelleyComparisonFourVowel2020;
@mulikVowelSystemSantiago2021;@stanleySampleSizeMatters2023;
@smithModelledMultivariateOverlap2024;@bernardiBhattacharyyasAffinityVs2025]. 

Now the question is: what do these measures actually do? One of the major problems is that a lot of researchers claim that they both show evidence that the vowels are overlapping [e.g., @nyczBestPracticesMeasuring2014 and Pillai-Bartlett ]. This is not actually true because Pillai-Bartlett Traces are the output of a MANOVA statistic and, like ANOVAs, explains whether or not the variance in the distributions are similar [@pillaiNewTestCriteria1955;@hardleAppliedMultivariateStatistical2024].[^1] The Pillai-Bartlett trace is only explaining is whether the two distributions are _similar_ to each other and **does not** actually explain overlap of the distributions.

On the otherhand, BA actually does explain the amount of overlap between the distributions [@bhattacharyyaMeasureDivergenceTwo1943]. This has been used quite extensively in ecological research to model the home ranges of various animals [@fiebergQuantifyingHomerangeOverlap2005;@seamanEffectsSampleSize1999]. It wasn't unitl @johnsonQuantifyingOverlapBhattacharyyas2015 that the linguistic world was introduced to BAs for vowel overlap. 

One thing that is nice about these two measures is that they both are typically bound on one end by 0 and 1 on the other.[^2] For BA, the closer the value is to 1 the higher the overlap. For Pillai, the closer it is to 0 the higher the similarity. 

Using these two measures should help explain the differences or lack thereof between between these three vowels. 

## Data description and exploration

In keeping with good scientific practices, we first need to do an exploratory analysis before we make our hypotheses. We can do this by looking at the overall shape and distributions of the raw data before and after we clean it. 

### Data Description

```{r}
#| label: load_data
#| message: false

bolognese <- readr::read_csv(
  here::here(
    "data",
    "raw",
    "20260904_master_cleaned.csv"
  ), 
  na = c("", NA),
  col_types = list(
    phoneme = "c",
    stress = "l",
    word = "f",
    time = "n",
    duration = "n",
    nextSeg = "f",
    F1 = "n",
    F2 = "n",
    F3 = "n",
    F4 = "n",
    speaker = "f", 
    lesson = "f"
  )
) |> 
  dplyr::mutate(
    environment = dplyr::case_when(
      nextSeg %in% c("ʎ", "l", "lː") ~ "prelateral", 
      nextSeg %in% c("m", "mː", "n", "nː", "ɲ", "ɲː", "ŋ", "ŋː") ~ "prenasal",
      nextSeg %in% c("rː", "r") ~ "prerhotic",
      TRUE ~ "elsewhere"
    )
  )
```

Data comes from the audio accompanying the lessons found in @vitaliDscdegarretBulgnaisManuale2009. Each lesson's audio file was transcribed, segmented, and annotated with which vowels were stressed and who the speaker was. Praat was used to calculate the duration of the segment and the first four resonant frequencies and was used to create a CSV file containing the phoneme, stress, word, where in the audio file it was located, and what the next segment was.

I then created a new variable indicating the phonological environment that the segments were found in: 

- `prelateral` if the segment occured before one of the laterals; 
- `prenasal` if the segment occured before any nasal consonant;
- `prerhotic` if the next segment was a rhotic; 
- `elsewhere` for everything else. These levels were selected because it is well known that laterals, nasals, and rhotics can greatly influence the vowels next to them via a phenomenon known as coarticulation. 

This gives us a dataframe with the following variables. 

```{r}
#| echo: false
#| label: tab_variables
#| tbl-cap: Description of the variables in the dataset

data_dictionary <- tibble::tibble(
  variable = names(bolognese), 
  type = purrr::map_chr(bolognese, ~class(.x)[1])
) |> 
  dplyr::mutate(
    description = c(
      "The phoneme", 
      "The syllable stress the segment is located in",
      "The word that the phoneme was found in",
      "THe time in seconds where the phoneme begins",
      "The duration of the phoneme",
      "The next segment in the word is",
      "The first resonant frequency in Hz",
      "The second resonant frequency in Hz",
      "The third resonant frequency in Hz",
      "The fourth resonant frequency in Hz",
      "The speakers initials",
      "The lesson number",
      "The phonological enivornment of the phoneme"
    )
  )

data_dictionary |> 
  knitr::kable(
    col.names = c("Variable", "Type", "Description")
  )
```

Now lets take a look at the distribution of vowels. Now we aren't going to be looking at all of the vowels because we are primarily interested in stressed monophthong vowels.

```{r}
#| label: vowel_splitting

mono <- c(
        "i",
        "iː",
        "ɛː",
        "e",
        "eː",
        "u",
        "uː",
        "ɔː",
        "o",
        "oː",
        "ɑ",
        "a",
        "aː",
        "æ"
      )

long <- c(
        "iː",
        "ɛː",
        "eː",
        "uː",
        "ɔː",
        "oː",
        "aː"
      )

vwls_stress <- bolognese |> 
  dplyr::filter(
    stress == TRUE,
    phoneme %in% mono
  ) |> 
  dplyr::mutate(
    vowel_length = dplyr::case_when(
      phoneme %in% long ~ "long",
      TRUE ~ "short"
    ),
    vowel_quality = dplyr::case_when(
      phoneme %in% c("i", "iː") ~ "i",
      phoneme %in% c("a", "aː") ~ "a",
      phoneme %in% c("e", "eː") ~ "e",
      phoneme %in% c("o", "oː") ~ "o",
      phoneme %in% c("u", "uː") ~ "u",
      TRUE ~ phoneme
    )
  )

vowel_count <- table(vwls_stress$phoneme)

vowel_environment <- table(vwls_stress$phoneme, vwls_stress$environment)

vowel_speaker <- table(vwls_stress$phoneme, vwls_stress$speaker)
```

::: {.panel-tabset}
## Stressed Vowels from dataset
```{r}
#| echo: false
#| label: phoneme_freq

vowel_count |> 
  knitr::kable(
    col.names = c("Phoneme", "Freq")
  )
```

## Stressed Vowels by environment
```{r}
#| echo: false
#| label: phoneme_environment

vowel_environment |> 
  knitr::kable(
    col.names = c("Phoneme", "Elsewhere", "Prelateral", "Prenasal", "Prerhotic")
  )
```

## Stressed Vowels by speaker
```{r}
#| echo: false
#| label: phoneme_speakers

vowel_speaker |> 
  knitr::kable(
    col.names = c("Phoneme", "LL", "RM")
  )
```
:::

From here it is good practice to remove outliers from the dataset. I find taking 95% of the data based on the mahalnobis distance really effective and luckily there is a handy way to do this using the `joeyr` package, which uses tidy syntax to calculate the mahalanobis distance and then to slowly delete rows until 95% of the data remains for the group variables. In this case, we will group by phoneme, speaker, and environment and keep 95% of those groups. 

```{r}
#| label: outlier_removal

vwls_clean <- vwls_stress |> 
  dplyr::mutate(
    is_outlier = joeyr::find_outliers(F1, F2, keep = 0.95),
    .by = c("phoneme", "speaker", "environment")
  ) |> 
  dplyr::filter(
    !is_outlier
  )
```

By doing this we only lost `{r} nrow(vwls_stress)-nrow(vwls_clean)` vowels from the original `{r} nrow(vwls_stress)`.

I also typically check the fundamental frequency (f0) and Energy of the signal since this impacts if Praat or Voicesauce was able to calculate the formants and other voice quality measures correctly. In those cases I z-score the f0 and only keep tokens within 3 SD of the mean. I also keep all tokens whose energy does not equal zero. I was not able to do that because I do not have those measures. 

### Vowel Plots

The next thing we need to do is normalize the vowels. There are a lot of different ways that we can do this but my favorite has to be the ∆F method described in @johnsonDFMethodVocal2020. This calculates a measure called $\Delta F$ which is the average formant difference and then divides the formant by this measure. This essentialy normalizes the formants by assuming everyone has the same vocal tract length. 

Once we have these new values we are ready to begin plotting. Luckily there is a very handy package called `tidynorm` which can do this for us.

```{r}
#| label: vowel_normalize

vwls_norm <- vwls_clean |> 
  tidynorm::norm_deltaF(
    F1:F4, 
    .by = speaker
  )
```

We are going to plot several different vowel plots. One based on vowel length, one by speaker, and one for each environment.


::: {.panel-tabset}

## Vowel length

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig_vowelplot-length


vwls_norm |>
  ggplot2::ggplot(aes(
    x = F2_df,
    y = F1_df,
    color = vowel_quality
  )) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::stat_ellipse(
    level = 0.67,
    linewidth = 1
  ) +
  ggplot2::scale_x_reverse(position = "top") +
  ggplot2::scale_y_reverse(position = "right") +
  ggokabeito::scale_colour_okabe_ito() +
  ggplot2::facet_wrap(. ~ vowel_length) + 
  ggplot2::labs(
    title = "∆F Normalized Vowels ",
    x = "F2 (∆F)",
    y = "F1 (∆F)"
  ) +
  ggplot2::theme_bw()
```

## Vowels by speaker

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig_vowelplot-speaker

vwls_norm |>
  ggplot2::ggplot(aes(
    x = F2_df,
    y = F1_df,
    color = vowel_quality
  )) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::stat_ellipse(
    level = 0.67,
    linewidth = 1
  ) +
  ggplot2::scale_x_reverse(position = "top") +
  ggplot2::scale_y_reverse(position = "right") +
  ggokabeito::scale_colour_okabe_ito() +
  ggplot2::facet_wrap(. ~ speaker) + 
  ggplot2::labs(
    title = "∆F Normalized Vowels ",
    x = "F2 (∆F)",
    y = "F1 (∆F)"
  ) +
  ggplot2::theme_bw()
```

## Vowels in prelateral

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig_vowelplot-prelateral

vwls_norm |>
  dplyr::filter(
    environment == "prelateral"
  ) |> 
  ggplot2::ggplot(aes(
    x = F2_df,
    y = F1_df,
    color = vowel_quality
  )) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::stat_ellipse(
    level = 0.67,
    linewidth = 1
  ) +
  ggplot2::scale_x_reverse(position = "top") +
  ggplot2::scale_y_reverse(position = "right") +
  ggokabeito::scale_colour_okabe_ito() +
  ggplot2::facet_wrap(. ~ vowel_length) + 
  ggplot2::labs(
    title = "Prelateral Vowels (∆F Normalized)",
    x = "F2 (∆F)",
    y = "F1 (∆F)"
  ) +
  ggplot2::theme_bw()
```

## Vowels in prenasal

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig_vowelplot-prenasal


vwls_norm |>
  dplyr::filter(
    environment == "prenasal"
  ) |> 
  ggplot2::ggplot(aes(
    x = F2_df,
    y = F1_df,
    color = vowel_quality
  )) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::stat_ellipse(
    level = 0.67,
    linewidth = 1
  ) +
  ggplot2::scale_x_reverse(position = "top") +
  ggplot2::scale_y_reverse(position = "right") +
  ggokabeito::scale_colour_okabe_ito() +
  ggplot2::facet_wrap(. ~ vowel_length) + 
  ggplot2::labs(
    title = "Prenasal Vowels (∆F Normalized)",
    x = "F2 (∆F)",
    y = "F1 (∆F)"
  ) +
  ggplot2::theme_bw()
```

## Vowels in prerhotic

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig_vowelplot-prerhotic

vwls_norm |>
  dplyr::filter(
    environment == "prerhotic"
  ) |> 
  ggplot2::ggplot(aes(
    x = F2_df,
    y = F1_df,
    color = vowel_quality
  )) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::stat_ellipse(
    level = 0.67,
    linewidth = 1
  ) +
  ggplot2::scale_x_reverse(position = "top") +
  ggplot2::scale_y_reverse(position = "right") +
  ggokabeito::scale_colour_okabe_ito() +
  ggplot2::facet_wrap(. ~ vowel_length) + 
  ggplot2::labs(
    title = "Prerhotic Vowels (∆F Normalized)",
    x = "F2 (∆F)",
    y = "F1 (∆F)"
  ) +
  ggplot2::theme_bw()
```

## Elsewhere Vowels

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig_vowelplot-elsewhere

vwls_norm |>
  dplyr::filter(
    environment == "elsewhere"
  ) |> 
  ggplot2::ggplot(aes(
    x = F2_df,
    y = F1_df,
    color = vowel_quality
  )) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::stat_ellipse(
    level = 0.67,
    linewidth = 1
  ) +
  ggplot2::scale_x_reverse(position = "top") +
  ggplot2::scale_y_reverse(position = "right") +
  ggokabeito::scale_colour_okabe_ito() +
  ggplot2::facet_wrap(. ~ vowel_length) + 
  ggplot2::labs(
    title = "Elsewhere Vowels (∆F Normalized)",
    x = "F2 (∆F)",
    y = "F1 (∆F)"
  ) +
  ggplot2::theme_bw()
```

:::

The plotting reveals some interesting behaviors. 

- There is a difference between how LL and RM produce the low vowels. RM seems to have them all overlap whereas LL doesn't. 
- The /æ/ vowel has two distributions. Prelateral /æ/ is closer to /e/ otherwise it appears low and front
- All three vowels are clustering together prenasaly and appear to have a large degree of overlap
- There are very few instances of low vowels prerhoticly. This might be a simple gap in the data or a general feature of the language


## Similarity and Overlap analysis

The analysis will consist of finding the pillai trace and BA for each pairwise comparison between the low vowels followed by a bootstrap analysis for determining the robustness of the measures. 

Each measure also has a treshold measure that indicates whether the distributions are similar based on @stanleySampleSizeMatters2023 and whether the distributions are overlapped based on @bernardiBhattacharyyasAffinityVs2025. 

If the Pillai trace is smaller than the threshold then this is evidence that vowels are similar. If the BA is higher than the threshold then this indicates that the distributions are overlapping.

I am not aware of linguists using bootstrapping to asses the robustness of the measures so I need to establish some rules for interpreting this. 

For Pillai Traces:

- If the entire range of the Bootstrap 95% CI lies below the threshold, this is evidence that these distributions are similar. 
- If the entire rannge of the Bootstrap 95% CI lies above the threshold, this is evidence that these distributions are not similar.
- If the Bootstrap 95% CI stradles the threshold, we do not have evidence for similarity.

For BA:

- If the entire range of the Bootstrap 95% CI lies below the threshold, this is evidence that these distributions are not overlapped. 
- If the entire rannge of the Bootstrap 95% CI lies above the threshold, this is evidence that these distributions are overlapped.
- If the Bootstrap 95% CI stradles the threshold, we do not have evidence for overlap. 

In order to assess the three low vowels we need to split the data into subsets based on speaker and the environments. We also will be removing prelateral vowels from the analysis because /æ/ is realized as [e]. This, however, is strong evidence that phonologically /æ/ is a seperate sound, that what is being orthographically represented as /æ/ is a seperate sound, or that this is a lexical effect from the small class of sounds that have /æ/ prelaterally.


```{r}
#| label: data_subsetting

### vowels 
frntlss <- vwls_norm |> 
  dplyr::filter(
    phoneme %in% c("a", "ɑ"),
    environment != "prelateral"
  )

bcklss <- vwls_norm |> 
  dplyr::filter(
    phoneme %in% c("æ", "a"),
    environment != "prelateral"
  )

cntrlss <- vwls_norm |> 
  dplyr::filter(
    phoneme %in% c( "æ", "ɑ"),
    environment != "prelateral"
  )

### Speaker subsetting
ll_vwls_bcklss <- vwls_norm |> 
  dplyr::filter(
    speaker == "LL",
    phoneme %in% c("æ", "a"),
    environment != "prelateral"
  )

ll_vwls_frntlss <- vwls_norm |> 
  dplyr::filter(
    speaker == "LL",
    phoneme %in% c("a", "ɑ"),
    environment != "prelateral"
  )

ll_vwls_cntrlss <- vwls_norm |> 
  dplyr::filter(
    speaker == "LL",
    phoneme %in% c("æ", "ɑ"),
    environment != "prelateral"
  )

rm_vwls_bcklss <- vwls_norm |> 
  dplyr::filter(
    speaker != "LL",
    phoneme %in% c("æ", "a"),
    environment != "prelateral"
  )

rm_vwls_frntlss <- vwls_norm |> 
  dplyr::filter(
    speaker != "rm",
    phoneme %in% c("a", "ɑ"),
    environment != "prelateral"
  )

rm_vwls_cntrlss <- vwls_norm |> 
  dplyr::filter(
    speaker != "LL",
    phoneme %in% c("æ", "ɑ"),
    environment != "prelateral"
  )
```


### Calculating the Pillai Trace 

Calculating the Pillai Trace is fairly straightforward and involves using the built in `manova` function in R. A MANOVA is the same thing as an ANOVA but uses multivariant data instead. The multivariate approach is better because vowels are inherently multivarient because they are based their position in space based on F1 and F2. 

In our analysis we will be using F1, F2 in our MANOVA. @kelleyComparisonFourVowel2020 found that adding duration as a third variable was more informative than F1 and F2 alone but given that there is some rather extreme behavior in duration due to stress we will not consider adding duration. 

```{r}
#| label: overall_pillai
#| warning: false

pillai_fn <- function(data, i) {
  
  df <- data[i,]

  overlap(F1 = df$F1_df[i], F2 = df$F2_df[i], vowel = df$phoneme[i], method = "pillai")
}

# frntlss
set.seed(1234)
boot_pillai <- boot::boot(frntlss, pillai_fn, R = 10000)
ci_pillai <- boot::boot.ci(boot_pillai, type = "bca")

frntless_pillai <- tibble::tibble(
  pillai = overlap(F1 = frntlss$F1_df, F2 = frntlss$F2_df, vowel = frntlss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(frntlss) / 2),
  # BCa_CI = [0.0302, 0.226]
)

#bcklss
set.seed(1234)
bcklss_boot <- boot::boot(bcklss, pillai_fn, R = 10000)
bcklss_ci <- boot::boot.ci(bcklss_boot, type = "bca")

bcklss_pillai <- tibble::tibble(
  pillai = overlap(F1 = bcklss$F1_df, F2 = bcklss$F2_df, vowel = bcklss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(bcklss) / 2),
  # BCa_CI = [0.0302, 0.226]
)

#cntrlss
set.seed(1234)
cntrlss_boot <- boot::boot(cntrlss, pillai_fn, R = 10000)
cntrlss_ci <- boot::boot.ci(cntrlss_boot, type = "bca")

cntrlss_pillai <- tibble::tibble(
  pillai = overlap(F1 = cntrlss$F1_df, F2 = cntrlss$F2_df, vowel = cntrlss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(cntrlss) / 2),
  # BCa_CI = [0.0302, 0.226]
)
```

```{r}
#| label: ll_pillai
#| warning: false

# ll frntlss
set.seed(1234)
ll_frntlss_boot <- boot::boot(ll_vwls_frntlss, pillai_fn, R = 10000)
ll_frntlss_ci <- boot::boot.ci(ll_frntlss_boot, type = "bca")

ll_frntlss_pillai <- tibble::tibble(
  pillai = overlap(F1 = ll_vwls_frntlss$F1_df, F2 = ll_vwls_frntlss$F2_df, vowel = ll_vwls_frntlss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(ll_vwls_frntlss) / 2),
  # BCa_CI = [0.0302, 0.226]
)

# ll bcklss
set.seed(1234)
ll_bcklss_boot <- boot::boot(ll_vwls_bcklss, pillai_fn, R = 10000)
ll_bcklss_ci <- boot::boot.ci(ll_bcklss_boot, type = "bca")

ll_bcklss_pillai <- tibble::tibble(
  pillai = overlap(F1 = ll_vwls_bcklss$F1_df, F2 = ll_vwls_bcklss$F2_df, vowel = ll_vwls_bcklss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(ll_vwls_bcklss) / 2),
  # BCa_CI = [0.0302, 0.226]
)

# ll cntrlss
set.seed(1234)
ll_cntrlss_boot <- boot::boot(ll_vwls_cntrlss, pillai_fn, R = 10000)
ll_cntrlss_ci <- boot::boot.ci(ll_cntrlss_boot, type = "bca")

ll_cntrlss_pillai <- tibble::tibble(
  pillai = overlap(F1 = ll_vwls_cntrlss$F1_df, F2 = ll_vwls_cntrlss$F2_df, vowel = ll_vwls_cntrlss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(ll_vwls_cntrlss) / 2),
  # BCa_CI = [0.0302, 0.226]
)
```


```{r}
#| label: rm_pillai
#| message: false
#| warning: false

# rm frntlss
set.seed(1234)
rm_frntlss_boot <- boot::boot(rm_vwls_frntlss, pillai_fn, R = 10000)
rm_frntlss_ci <- boot::boot.ci(rm_frntlss_boot, type = "bca")

rm_frntlss_pillai <- tibble::tibble(
  pillai = overlap(F1 = rm_vwls_frntlss$F1_df, F2 = rm_vwls_frntlss$F2_df, vowel = rm_vwls_frntlss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(rm_vwls_frntlss) / 2),
  # BCa_CI = [0.0302, 0.226]
)

# rm bcklss
set.seed(1234)
rm_bcklss_boot <- boot::boot(rm_vwls_bcklss, pillai_fn, R = 10000)
rm_bcklss_ci <- boot::boot.ci(rm_bcklss_boot, type = "bca")

rm_bcklss_pillai <- tibble::tibble(
  pillai = overlap(F1 = rm_vwls_bcklss$F1_df, F2 = rm_vwls_bcklss$F2_df, vowel = rm_vwls_bcklss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(rm_vwls_bcklss) / 2),
  # BCa_CI = [0.0302, 0.226]
)

# rm cntrlss
set.seed(1234)
rm_cntrlss_boot <- boot::boot(rm_vwls_cntrlss, pillai_fn, R = 10000)
rm_cntrlss_ci <- boot::boot.ci(rm_cntrlss_boot, type = "bca")

rm_cntrlss_pillai <- tibble::tibble(
  pillai = overlap(F1 = rm_vwls_cntrlss$F1_df, F2 = rm_vwls_cntrlss$F2_df, vowel = rm_vwls_cntrlss$phoneme, method = "pillai"),
  pillai_cutoff = exp(1) / (nrow(rm_vwls_cntrlss) / 2),
  # BCa_CI = [0.0302, 0.226]
)
```

| Comparison | Pillai | Pillai cutoff | BCa 95% CI | 
|:-----------|:------:|:-------------:|:----------:|
| æ vs. a    | `{r} bcklss_pillai$pillai` | `{r} bcklss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(bcklss_ci)$bca`] |
| a vs. ɑ    | `{r} frntless_pillai$pillai` | `{r} frntless_pillai$pillai_cutoff` | [`{r} fwb::get_ci(ci_pillai)$bca`] |
| æ vs. ɑ    | `{r} cntrlss_pillai$pillai` | `{r} cntrlss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(cntrlss_ci)$bca`] |
| LL's æ vs. a | `{r} ll_bcklss_pillai$pillai` | `{r} ll_bcklss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(ll_bcklss_ci)$bca`] |
| LL's a vs. ɑ | `{r} ll_frntlss_pillai$pillai` | `{r} ll_frntlss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(ll_frntlss_ci)$bca`] |
| LL's æ vs. ɑ | `{r} ll_cntrlss_pillai$pillai` | `{r} ll_cntrlss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(ll_cntrlss_ci)$bca`] |
| RM's æ vs. a | `{r} rm_bcklss_pillai$pillai` | `{r} rm_bcklss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(rm_bcklss_ci)$bca`] |
| RM's a vs. ɑ | `{r} rm_frntlss_pillai$pillai` | `{r} rm_frntlss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(rm_frntlss_ci)$bca`] |
| RM's æ vs. ɑ | `{r} rm_cntrlss_pillai$pillai` | `{r} rm_cntrlss_pillai$pillai_cutoff` | [`{r} fwb::get_ci(rm_cntrlss_ci)$bca`] |

### Calculating the Bhattacharrya's Affinity


```{r}
#| label: overall_ba
#| warning: false

ba_fn <- function(data, i) {
  
  df <- data[i,]

  overlap(F1 = df$F1_df[i], F2 = df$F2_df[i], vowel = df$phoneme[i], method = "BA")
}

# frntlss
set.seed(1234)
boot_ba <- boot::boot(frntlss, ba_fn, R = 10000)
ci_ba <- boot::boot.ci(boot_ba, type = "bca")

frntless_ba <- tibble::tibble(
  ba = overlap(F1 = frntlss$F1_df, F2 = frntlss$F2_df, vowel = frntlss$phoneme, method = "BA"),
  cor = (cor_formant(frntlss, F1 = F1_df, F2 = F2_df)),
  ex = expected_BA(frntlss, vowel1 = "a", vowel2 = "ɑ"),
  ba_cutoff = threshold_BA(cor, ex)
  # BCa_CI = [0.0302, 0.226]
)

#bcklss
set.seed(1234)
bcklss_boot_ba <- boot::boot(bcklss, ba_fn, R = 10000)
bcklss_ci_ba <- boot::boot.ci(bcklss_boot_ba, type = "bca")

bcklss_ba <- tibble::tibble(
  ba = overlap(F1 = bcklss$F1_df, F2 = bcklss$F2_df, vowel = bcklss$phoneme, method = "BA"),
  cor = (cor_formant(bcklss, F1 = F1_df, F2 = F2_df)),
  ex = expected_BA(bcklss, vowel1 = "æ", vowel2 = "a"),
  ba_cutoff = threshold_BA(cor, ex)
  # BCa_CI = [0.0302, 0.226]
)

#cntrlss
set.seed(1234)
cntrlss_boot_ba <- boot::boot(cntrlss, ba_fn, R = 10000)
cntrlss_ci_ba <- boot::boot.ci(cntrlss_boot_ba, type = "bca")

cntrlss_ba <- tibble::tibble(
  ba = overlap(F1 = cntrlss$F1_df, F2 = cntrlss$F2_df, vowel = cntrlss$phoneme, method = "BA"),
  cor = (cor_formant(cntrlss, F1 = F1_df, F2 = F2_df)),
  ex = expected_BA(cntrlss, vowel1 = "æ", vowel2 = "ɑ"),
  ba_cutoff = threshold_BA(cor, ex)
  # BCa_CI = [0.0302, 0.226]
)
```

```{r}
#| label: ll_ba
#| warning: false

# ll frntlss
set.seed(1234)
ll_frntlss_boot_ba <- boot::boot(ll_vwls_frntlss, ba_fn, R = 10000)
ll_frntlss_ci_ba <- boot::boot.ci(ll_frntlss_boot_ba, type = "bca")

ll_frntlss_ba <- tibble::tibble(
  ba = overlap(F1 = ll_vwls_frntlss$F1_df, F2 = ll_vwls_frntlss$F2_df, vowel = ll_vwls_frntlss$phoneme, method = "BA"),
  cor = (cor_formant(ll_vwls_frntlss, F1 = F1_df, F2 = F2_df)),
  ex = expected_BA(ll_vwls_frntlss, vowel1 = "a", vowel2 = "ɑ"),
  ba_cutoff = threshold_BA(cor, ex)
)

# ll bcklss
set.seed(1234)
ll_bcklss_boot_ba <- boot::boot(ll_vwls_bcklss, ba_fn, R = 10000)
ll_bcklss_ci_ba <- boot::boot.ci(ll_bcklss_boot_ba, type = "bca")

ll_bcklss_ba <- tibble::tibble(
  ba = overlap(F1 = ll_vwls_bcklss$F1_df, F2 = ll_vwls_bcklss$F2_df, vowel = ll_vwls_bcklss$phoneme, method = "BA"),
  cor = (cor_formant(ll_vwls_bcklss, F1 = F1_df, F2 = F2_df)),
  ex = expected_BA(ll_vwls_bcklss, vowel1 = "æ", vowel2 = "a"),
  ba_cutoff = threshold_BA(cor, ex)
)

# ll cntrlss
set.seed(1234)
ll_cntrlss_boot_ba <- boot::boot(ll_vwls_cntrlss, ba_fn, R = 10000)
ll_cntrlss_ci_ba <- boot::boot.ci(ll_cntrlss_boot_ba, type = "bca")

ll_cntrlss_ba <- tibble::tibble(
  ba = overlap(F1 = ll_vwls_cntrlss$F1_df, F2 = ll_vwls_cntrlss$F2_df, vowel = ll_vwls_cntrlss$phoneme, method = "BA"),
  cor = (cor_formant(ll_vwls_cntrlss, F1 = F1_df, F2 = F2_df)),
  ex = expected_BA(ll_vwls_cntrlss, vowel1 = "æ", vowel2 = "ɑ"),
  ba_cutoff = threshold_BA(cor, ex)
)
```

```{r}
#| label: rm_ba
#| message: false
#| warning: false

# rm frntlss
set.seed(1234)
rm_frntlss_boot_ba <- boot::boot(rm_vwls_frntlss, ba_fn, R = 10000)
rm_frntlss_ci_ba <- boot::boot.ci(rm_frntlss_boot_ba, type = "bca")

rm_frntlss_ba <- tibble::tibble(
  ba = overlap(F1 = rm_vwls_frntlss$F1_df, F2 = rm_vwls_frntlss$F2_df, vowel = rm_vwls_frntlss$phoneme, method = "BA"),
  cor = cor_formant(rm_vwls_frntlss, F1 = F1_df, F2 = F2_df),
  ex = expected_BA(rm_vwls_frntlss, vowel1 = "a", vowel2 = "ɑ"),
  ba_cutoff = threshold_BA(cor, ex)
)

# rm bcklss
set.seed(1234)
rm_bcklss_boot_ba <- boot::boot(rm_vwls_bcklss, ba_fn, R = 10000)
rm_bcklss_ci_ba <- boot::boot.ci(rm_bcklss_boot_ba, type = "bca")

rm_bcklss_ba <- tibble::tibble(
  ba = overlap(F1 = rm_vwls_bcklss$F1_df, F2 = rm_vwls_bcklss$F2_df, vowel = rm_vwls_bcklss$phoneme, method = "BA"),
  cor = cor_formant(rm_vwls_bcklss, F1 = F1_df, F2 = F2_df),
  ex = expected_BA(rm_vwls_bcklss, vowel1 = "æ", vowel2 = "a"),
  ba_cutoff = threshold_BA(cor, ex)
)

# rm cntrlss
set.seed(1234)
rm_cntrlss_boot_ba <- boot::boot(rm_vwls_cntrlss, ba_fn, R = 10000)
rm_cntrlss_ci_ba <- boot::boot.ci(rm_cntrlss_boot_ba, type = "bca")

rm_cntrlss_ba <- tibble::tibble(
  ba = overlap(F1 = rm_vwls_cntrlss$F1_df, F2 = rm_vwls_cntrlss$F2_df, vowel = rm_vwls_cntrlss$phoneme, method = "BA"),
  cor = cor_formant(rm_vwls_cntrlss, F1 = F1_df, F2 = F2_df),
  ex = expected_BA(rm_vwls_cntrlss, vowel1 = "æ", vowel2 = "ɑ"),
  ba_cutoff = threshold_BA(cor, ex)
)
```

| Comparison | BA | BA cutoff | BCa 95% CI | 
|:-----------|:------:|:-------------:|:----------:|
| æ vs. a    | `{r} bcklss_ba$ba` | `{r} bcklss_ba$ba_cutoff` | [`{r} fwb::get_ci(bcklss_ci_ba)$bca`] |
| a vs. ɑ    | `{r} frntless_ba$ba` | `{r} frntless_ba$ba_cutoff` | [`{r} fwb::get_ci(ci_ba)$bca`] |
| æ vs. ɑ    | `{r} cntrlss_ba$ba` | `{r} cntrlss_ba$ba_cutoff` | [`{r} fwb::get_ci(cntrlss_ci_ba)$bca`] |
| LL's æ vs. a | `{r} ll_bcklss_ba$ba` | `{r} ll_bcklss_ba$ba_cutoff` | [`{r} fwb::get_ci(ll_bcklss_ci_ba)$bca`] |
| LL's a vs. ɑ | `{r} ll_frntlss_ba$ba` | `{r} ll_frntlss_ba$ba_cutoff` | [`{r} fwb::get_ci(ll_frntlss_ci_ba)$bca`] |
| LL's æ vs. ɑ | `{r} ll_cntrlss_ba$ba` | `{r} ll_cntrlss_ba$ba_cutoff` | [`{r} fwb::get_ci(ll_cntrlss_ci_ba)$bca`] |
| RM's æ vs. a | `{r} rm_bcklss_ba$ba` | `{r} rm_bcklss_ba$ba_cutoff` | [`{r} fwb::get_ci(rm_bcklss_ci_ba)$bca`] |
| RM's a vs. ɑ | `{r} rm_frntlss_ba$ba` | `{r} rm_frntlss_ba$ba_cutoff` | [`{r} fwb::get_ci(rm_frntlss_ci_ba)$bca`] |
| RM's æ vs. ɑ | `{r} rm_cntrlss_ba$ba` | `{r} rm_cntrlss_ba$ba_cutoff` | [`{r} fwb::get_ci(rm_cntrlss_ci_ba)$bca`] |

### Overall patterns

- Overall æ ~ a shows a lack of similarity and that the distributions are not overlapped.
- Overall a ~ ɑ shows a lack of similarity and the distributions are not overlapped.
- Overall æ ~ ɑ is uncertain when it comes to similarity and uncertainty about overlap.
- LL's æ ~ a shows a lack of similarity and a lack of overlap
- LL's a ~ ɑ shows a lack of similarity and a lack of overlap
- LL's æ ~ ɑ shows a lack of similarity and a lack overlap
- RM's æ ~ a shows an uncertainty of similarity and uncertainty about overlap in distributions
- RM's a ~ ɑ shows a lack of similarity and lack of overlap in distributions
- RM's æ ~ ɑ shows that the vowels are similar and overlap in distributions

My opinion is that the three vowels are separate phonemes except for RM's æ ~ ɑ pair which shows a high degree of similarity and overlap in distributions. Additionally, RM's æ ~ a vowels show some uncertainty about similarity and overlap.

## References

::: {#refs}
:::

[^1]: Two fantastic resources going into all of this are @hastieElementsStatisticalLearning2009 and @jamesIntroductionStatisticalLearning2021. 

[^2]: This is only true when we are doing pairwise comparisons. The upper bound of a Pillai is actually $k-1$, where $k$ is the number of groups that are being compared [@prabhakaranMANOVATestMultiple2026]. Since linguists are often doing pairwise comparisons this is effectively bound at 1, since $2-1=1$.

